Human Factors Influencing Visual Statistical Inference

### 2013-06-06 sociology_paper.tex
revising intro

In visual inference, the test statistic is a plot of the observed data. To create a lineup, this plot, called actual data plot, is placed on a layout of null plots. The null plots  are generated from the model specified by a null hypothesis, essentially describing what the plot might look like if the data had no structure. An observer is asked to evaluate the lineup. If the  actual data plot is detected by the observer, the null hypothesis is rejected. This means that the structure in the actual data plot has significant structure, a pattern that is not simply due to randomness.

\subsection{Visual inference example}
 Figure~\ref{fig:tower} displays a lineup of 20 plots where one of the plots is observed data, while the remaining 19 plots are rendered from data generated under a null model. Which of these towers is the most extreme? Which side is higher? In an experiment 12 of 72 observers picked plot \# $3^2+4$ as the most extreme plot from the lineup, based on the highest amount of asymmetry (36\%), a trend (26\%) or `outliers' (16\%). 
This gives us a visual $p$-value of 0.00023, indicating that we have enough evidence to reject the null hypothesis (in another four repetitions we found similar results, giving an overall $p$-value of less that $10^{-17}$ based on 62 data picks out of 341 evaluation). 
What does this mean, though? For that, we need to know the context of the data and we need to have more information about the generation of the null plots: in this example, we are investigating, whether the results from the 2012 US presidential election is consistent with polling results. As a null hypothesis we  assume that  the election results are consistent with polling results. The polling results give us a null model to generate data. Each poll gives us a margin between candidates. This is what we use as the mean of a normal distribution. In the absence of information of error for a poll, we will assume a standard deviation of 2.5 percentage points for all of the polls. While this is a severe simplification, the resulting margin of error of $\pm 5$ percentage points is realistic, and could be easily replaced if additional information on a poll became available. A null data set is then generate as a set of draws from normal distributions based on the latest poll results. Each panel in figure~\ref{fig:tower} shows an `electoral building' \citet{nytimes}: each state is represented by a rectangle. The width of a rectangle is determined by the difference between (polling) results, its height is given by the number of electoral votes, and fill color shows party affiliation (blue for Democrat, and red for Republican). Rectangles are placed on top of each other, separated by affiliation and sorted by margin between polls. A party needs 270 electoral votes or more to win the presidential election.

@@

\subsection{Performance of the observer} Observer who evaluates a lineup do not necessarily be aware of the data that constitue a lineup. In response to a specific question the observer is to make a choice based on what he or she sees in the lineup. The choice could be right or wrong depending on whether the data plot is selected or not. Under null hypothesis, all the null plots look similar to what the observed data plot appears to be. This would make it harder to detect the observed data plot in the lineup. It is not expected that an observer would be able to detect the data plot in this scenario. But since there are limited number of plots in a lineup which is 20 in this case, there is a 1/20 chance that the observer would pick the actual plot. This proportion is associated with the type-I error probability. On the other hand if null hypothesis is not true, the observed plot would be different than the null plots making it easier to be distinguished. When multiple observers evaluate a lineup, the proportion of correct response can be used to estimate the power of the visual test \citep{majumder:2013}.

@@

>>>>

In visual inference, the test statistic is a plot of the observed data. To create a lineup, this plot, called actual data plot, is placed in a layout of null plots. The null plots  are generated from the model specified by a null hypothesis, essentially describing what the plot might look like if the data had no structure. An observer is asked to evaluate the lineup. If the  actual data plot is detected by the observer, the null hypothesis is rejected. This means that the structure in the actual data plot has significant structure, a pattern that is not simply due to randomness. Combining the choices of multiple observers provides more stability in the estimation of significance.

\subsection{Visual inference example}

Figure~\ref{fig:tower} displays a lineup of 20 plots where one of the plots is observed data, while the remaining 19 plots are rendered from data generated under a null model. Which of these towers is the most extreme? Which side is higher? In an experiment 12 of 72 observers picked plot \# $3^2+4$ as the most extreme plot from the lineup, based on the highest amount of asymmetry (36\%), a trend (26\%) or `outliers' (16\%). 
This gives us a visual $p$-value of 0.00023, indicating that we have enough evidence to reject the null hypothesis (in another four repetitions we found similar results, giving an overall $p$-value of less that $10^{-17}$ based on 62 data picks out of 341 evaluation). 
What does this mean, though? For that, we need to know the context of the data and we need to have more information about the generation of the null plots: in this example, we are investigating, whether the results from the 2012 US presidential election is consistent with polling results. As a null hypothesis we  assume that  the election results are consistent with polling results. The polling results give us a null model to generate data. Each poll gives us a margin between candidates. This is what we use as the mean of a normal distribution. In the absence of information of error for a poll, we will assume a standard deviation of 2.5 percentage points for all of the polls. While this is a severe simplification, the resulting margin of error of $\pm 5$ percentage points is realistic, and could be easily replaced if additional information on a poll became available. A null data set is then generate as a set of draws from normal distributions based on the latest poll results. Each panel in figure~\ref{fig:tower} shows an `electoral building' \citet{nytimes}: each state is represented by a rectangle. The width of a rectangle is determined by the difference between (polling) results, its height is given by the number of electoral votes, and fill color shows party affiliation (blue for Democrat, and red for Republican). Rectangles are placed on top of each other, separated by affiliation and sorted by margin between polls. A party needs 270 electoral votes or more to win the presidential election.

@@

\subsection{Observer task} Observer who evaluates a lineup do not necessarily be aware of the data that constitue a lineup. In response to a specific question the observer is to make a choice based on what he or she sees in the lineup. The choice could be right or wrong depending on whether the data plot is selected or not. Under null hypothesis, all the null plots look similar to what the observed data plot appears to be. This would make it harder to detect the observed data plot in the lineup. It is not expected that an observer would be able to detect the data plot in this scenario. But since there are limited number of plots in a lineup which is 20 in this case, there is a 1/20 chance that the observer would pick the actual plot. This proportion is associated with the type-I error probability. On the other hand if null hypothesis is not true, the observed plot would be different than the null plots making it easier to be distinguished. When multiple observers evaluate a lineup, the proportion of correct response can be used to estimate the power of the visual test \citep{majumder:2013}.

@@

\subsection{Experiments pooled}
